### 2016-04-07 ClimaTrans/climatrans.Rmd
Results for Climatrans: precipitation and hot summer days

**Analysis** 
This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. To execute this type:

@@

##*Precipitation*

#### Function used to estimate proportional change (for precipitation):

```{r,eval=FALSE}
propchange <- function(X,it0=c(1979,2013)) {
  z <- coredata(X)
  if (is.null(dim(z)))
      z <- 100*(z/mean(coredata(subset(X,it=it0)),na.rm=TRUE)) else
      z <- 100*t(t(z)/apply(coredata(subset(X,it=it0)),2,'mean',na.rm=TRUE))
  attributes(z) <- NULL
  z -> coredata(X)  
  X
}
```

Function that converts the surface temperature into saturation vapour pressure:

```{r}
tas2es <- function(x,mask=FALSE,land=TRUE,season=6:9,FUN='min') {
  if (mask) x <- mask(x,land=land)
  es <- subset(C.C.eq(x),it=month.abb[season])
  es <- aggregate(es,year,FUN=FUN)
  invisible(es)
}
```

@@

## Combine the individual stations into a group of stations
climatrans.pr <- combine(delhi,bombay,bangalore)
## Save the results for future use
save(file='~/Dropbox/Public/ClimaTrans/climatrans.pr.rda',climatrans.pr)

@@

## Combine the individual stations into a group of stations
climatrans.tx <- combine(tx.delhi,tx.bombay,tx.bangalore)
## Save the results for future use
save(file='~/Dropbox/Public/ClimaTrans/climatrans.tx.rda',climatrans.tx)
```

@@

### **Daily station data**
```{r,echo=TRUE}
## Load precipitation data extracted from the gridded IMD data set:
load('~/Dropbox/Public/ClimaTrans/climatrans.pr.rda')
## Display the structure of the data as
str(climatrans.pr)
```

@@

```{r, fig.width=6, fig.height=6}
## Aggregate to mean seasonal cycle and annual mean values
plot(aggregate(climatrans.pr,month,'sum'),new=FALSE)
```

Estimate annual aggregated statistics for the months May-October (the wet season):

```{r}
fw <- aggregate(subset(climatrans.pr,it=month.abb[5:10]),year,'wetfreq')
mu <- aggregate(subset(climatrans.pr,it=month.abb[5:10]),year,'wetmean')
q95 <- aggregate(subset(climatrans.pr,it=month.abb[5:10]),year,'quantile',probs=0.95)

## check the relationship between wet-day 95-percentile and wet-day mean
plot(zoo(mu[,2]),zoo(q95[,2]),pch=19,xlim=c(0,40),ylim=c(0,120),
     main='Wet-day mean v.s. 95th wet percentile',
     xlab=expression(paste(mu,' (mm/day)')), ylab=expression(paste(q[95],' (mm/day)')))
points(zoo(mu[,1]),zoo(q95[,1]),pch=19,col=rgb(1,0,0,0.2))
points(zoo(mu[,3]),zoo(q95[,3]),pch=19,col=rgb(0,0,1,0.2))
grid()
legend(0,120,loc(climatrans.pr),pch=19,col=c(rgb(1,0,0,0.2),'black',rgb(0,0,1,0.2)),bty='n')
```

@@

```{r}
## Aggregate sub-season values to annual mean values
if (!exists("es")) {
  #sst <- retrieve('data/ncep/sst.mon.mean.nc',lon=c(50,100),lat=c(0,40))
  #es <- aggregate(C.C.eq(subset(sst,it=month.abb[6:9])),year,'max')
  t2m <- retrieve('data/ncep/air.mon.mean.nc',lon=c(50,100),lat=c(5,30))
  es <- annual(t2m,'tas2es')
}
```

@@

```{r}
if (!exists("slp")) {
  slp <- retrieve('data/ncep/slp.mon.mean.nc',lon=c(50,100),lat=c(0,40))
  slp <- aggregate(subset(slp,it=month.abb[6:9]),year,'mean')
}

@@

### 
```{r,echo=FALSE, fig.width=6, fig.height=6}
## The function DS uses multiple regression by default
ds.mu <- DS(pca.mu,eof.es,eofs=1:20)
ds.fw <- DS(pca.fw,eof.slp,eofs=1:20)
plot(ds.mu,colbar1=list(breaks=seq(0,1,by=0.1),rev=2),
     colbar2=list(breaks=seq(-500,500,by=25),rev=TRUE))
plot(ds.fw)
```

#### **Compare the histogram of the preciptation data with an exponential distribution**
```{r}
z <- climatrans.pr[,3]; z <- z[z > 1]
hist(z,breaks=seq(0,500,by=2),xlim=c(0,100),freq=FALSE)
lines(seq(0,250,by=1),dexp(seq(0,250,by=1),rate=1/mean(z)),lwd=3,col='red')
```

@@

Here include the argument \code{xfuns='tas2es'} to use the option \code{annual(tas2es(gcm))} rather than the option \code{annual(gcm,FUN='tas2es')} which will fail. Also

```{r}
if (!file.exists('dse.mu.climatrans.rda')) {
  dse <- DSensemble(pca.mu,predictor=es,FUNX='tas2es',xfuns='tas2es',nmin=1)
  save(file='dse.mu.climatrans.rda',dse)
} else load('dse.mu.climatrans.rda')
dse.stations<-as.station(dse)
```

@@

Check the long-term change: historic trend in the mean

```{r}
  load('~/Dropbox/Public/ClimaTrans/climatrans.tx.rda')
  plot(as.4seasons(anomaly(climatrans.tx)))
```

Downscale the seasonal seasonal mean maximum temperature

```{r}
  predictor <- retrieve('data/ncep/air.mon.mean.nc',
                        param='air',lon=c(60,90),lat=c(10,35))

  dev.new()
  trh <- c(40,30,32)
  for (i in 1:3) {
    y <- subset(climatrans.tx,is=i)
    if (!file.exists(paste('dse.tx.climatrans.',loc(y),'.rda',sep=''))) {
      dse.tx <- DSensemble.t2m(y,biascorrect=TRUE,type='ncdf4',
                           predictor=predictor,nmin=60,verbose=FALSE)
      save(file=paste('dse.tx.climatrans.',loc(y),'.rda',sep=''),dse.tx)
    } else load(paste('dse.tx.climatrans.',loc(y),'.rda',sep=''))

    ## The function hotsummerdays estimates the number of hot days based on
    ## the assumption that the temperature is close to being normally distributed
    hw <- hotsummerdays(x=y,dse=dse.tx,threshold=trh[i],plot=FALSE)
    plot(hw)
  }
```

@@

Check the long-term change: historic trend in the seasonal spread (standard deviation)

```{r}
  load('~/Dropbox/Public/ClimaTrans/climatrans.tx.rda')
  plot(as.4seasons(anomaly(climatrans.tx),FUN='sd'))
```

>>>>

**Analysis** 
The downscaling of the wet-day mean precipitation used the surface temperature as predictor.

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. To execute this type:

@@

##*Functions*

Define a function that converts the surface temperature into saturation vapour pressure:

```{r}
tas2es <- function(x,mask=TRUE,land=TRUE,season=5:10,FUN='min') {
  if (mask) x <- mask(x,land=land)
  es <- subset(C.C.eq(x),it=month.abb[season])
  es <- aggregate(es,year,FUN=FUN)
  index(es) <- year(es)
  invisible(es)
}
```

Function to estimate probabilty for daily amount of exceeding a threshold x0 given that the distribution follows an exponential distribution and the records of the wet-day mean precipitation and frequency.

```{r}
PrexpPr <- function(mu,fw,x0=10) {
  Pr <- zoo(coredata(fw)*exp(-x0/coredata(mu)),order.by=year(fw))

  Pr.mu <- zoo(mean(coredata(fw))*exp(-x0/coredata(mu)),order.by=year(fw))
  Pr.fw <- zoo(coredata(fw)*exp(-x0/mean(coredata(mu))),order.by=year(fw))
  attr(Pr,'prob.f(mu)') <- Pr.mu
  attr(Pr,'prob.f(fw)') <- Pr.fw
  Pr <- attrcp(mu,Pr)
  attr(Pr,'variable') <- 'Pr'
  attr(Pr,'unit') <- 'probability'
  attr(Pr,'longname') <- paste('Probability of exceeding',x0)
  class(Pr) <- class(mu)
  invisible(Pr)
}
```
##*Precipitation*

@@

## Combine the individual stations into a group of stations
climatrans.pr <- combine(delhi,bombay,bangalore)
## Save the results for future use
save(file='climatrans.pr.rda',climatrans.pr)

@@

## Combine the individual stations into a group of stations
climatrans.tx <- combine(tx.delhi,tx.bombay,tx.bangalore)
## Save the results for future use
save(file='climatrans.tx.rda',climatrans.tx)
```

@@

### **Daily station data**
```{r,echo=TRUE}
## Load precipitation data extracted from the gridded IMD data set:
load('climatrans.pr.rda')
## Display the structure of the data as
str(climatrans.pr)
```
##**Sensitivity tests**

How do the precipitation respond to varying forcings/conditions? Such iformation may provide useful clues as to how the precipitation is expected to change in the future.

@@

```{r, fig.width=6, fig.height=6}
## Monthly precipitation totals
plot(aggregate(climatrans.pr,month,'sum'),new=FALSE)
```

Estimate annual aggregated statistics for the months May-October (the wet/Monsoon season):

```{r}
season<-5:10
fw <- aggregate(subset(climatrans.pr,it=month.abb[season]),year,'wetfreq')
mu <- aggregate(subset(climatrans.pr,it=month.abb[season]),year,'wetmean')
q95 <- aggregate(subset(climatrans.pr,it=month.abb[season]),year,'quantile',probs=0.95)
```

Compare the data to the exponential distribution - if the data is exponentially distributed, then the wet-day mean precipitation $\mu$ and the 95-percentile $q_{95}$ are propportional by $q_p = -ln(1-p)\mu$.

```{r}
## check the relationship between wet-day 95-percentile and wet-day mean
plot(-log(0.05)*zoo(mu[,2]),zoo(q95[,2]),pch=19,xlim=c(0,120),ylim=c(0,120),
     main='Wet-day mean v.s. 95th wet percentile',
     xlab=expression(paste(-ln(1-0.95)*mu,' (mm/day)')), ylab=expression(paste(q[95],' (mm/day)')))
points(-log(0.05)*zoo(mu[,1]),zoo(q95[,1]),pch=19,col=rgb(1,0,0,0.2))
points(-log(0.05)*zoo(mu[,3]),zoo(q95[,3]),pch=19,col=rgb(0,0,1,0.2))
grid()
legend(0,120,loc(climatrans.pr),pch=19,col=c(rgb(1,0,0,0.2),'black',rgb(0,0,1,0.2)),bty='n')
```

Compare the histogram of the preciptation data with an exponential distribution
```{r}
z <- climatrans.pr[,3]; z <- z[z > 1]
hist(z,breaks=seq(0,500,by=2),xlim=c(0,100),freq=FALSE)
lines(seq(0,250,by=1),dexp(seq(0,250,by=1),rate=1/mean(z)),lwd=3,col='red')
```

##**Historical long-term trends**

Estimate the long-term trends in the precipitation statistics at the three megacities: first the wet-day frequency $f_w$ - does the number of rainy days change over time?

```{r, fig.width=6, fig.height=6}
## Wet-day frequency
plot(fw,new=FALSE)
for(i in 1:3) lines(trend(subset(fw,is=i)))
```

Does the wet-day mean precipitation $\mu$ change over time?

```{r, fig.width=6, fig.height=6}
## Wet-day mean
plot(mu,new=FALSE)
for(i in 1:3) lines(trend(subset(mu,is=i)))
```

If the 24-hr preciptation amounts is exponential (when it rains), then the probability of exceeding a threshold is $Pr(X>x_0) = f_w e^{-x_0/\mu}$. Since we know both

```{r, fig.width=6, fig.height=6}
## Wet-day mean
Pr <- PrexpPr(mu,fw,x0=50) 
plot(Pr,new=FALSE,errorbar=FALSE)
for(i in 1:3) lines(trend(subset(Pr,is=i)))
```

Also, it's interesting to see how the annual total precipitation amount changes over time:

```{r, fig.width=6, fig.height=6}
## Aggregate to mean seasonal cycle and annual mean values
plot(annual(climatrans.pr,'sum'),new=FALSE)
for(i in 1:3) lines(trend(subset(annual(climatrans.pr,'wetfreq'),is=i)))
```

##**Predictors**

@@

```{r}
## Aggregate sub-season values to annual mean values
if (!exists("es")) {
  t2m <- retrieve('data/ncep/air.mon.mean.nc',lon=c(60,100),lat=c(0,23))
  es <- aggregate(tas2es(t2m,season=season),year,'mean')
}
```

@@

```{r}
if (!exists("slp")) {
  slp <- retrieve('data/ncep/slp.mon.mean.nc',lon=c(60,100),lat=c(0,23))
  slp <- aggregate(subset(slp,it=month.abb[season]),year,'mean')
}

@@

### 
```{r,echo=FALSE, fig.width=6, fig.height=6}
## The function DS uses multiple regression by default
ds.mu <- DS(pca.mu,eof.es,eofs=1:20)
ds.fw <- DS(pca.fw,eof.slp,eofs=1:20)
plot(ds.mu,colbar1=list(breaks=seq(0,1,by=0.1),rev=2),
     colbar2=list(breaks=seq(-500,500,by=25),rev=TRUE),new=FALSE)
plot(ds.fw,new=FALSE)
```

## **Examine the residuals from the downscaling**

First check the residuals for the wet-day mean precipitation:

```{r}
## Examine the PCA of residuals for the cold monthsto see if there are any structures left
  mu.2 <- as.residual(as.station(ds.mu))
  mu.2 <- attrcp(mu,mu.2); class(mu.2) <- class(mu)
  pca.mu.2 <- PCA(mu.2)
  plot(pca.mu.2)
```

The residuals for the wet-day frequency:

```{r}
## Examine the PCA of residuals for the cold monthsto see if there are any structures left
  fw.2 <- as.residual(as.station(ds.fw))
  fw.2 <- attrcp(fw,fw.2); class(fw.2) <- class(fw)
  pca.fw.2 <- PCA(fw.2)
  plot(pca.fw.2)
```

@@

Here include the argument xfuns='tas2es' to use the option annual(tas2es(gcm)) rather than the option annual(gcm,FUN='tas2es') which will fail. Also

```{r}
if (!file.exists('dse.mu.climatrans.rda')) {
  index(pca.mu) <- as.Date(paste(year(pca.mu),'01-01',sep='-'))
  index(pca.fw) <- as.Date(paste(year(pca.fw),'01-01',sep='-'))
  dse.mu <- DSensemble(pca.mu,predictor=es,FUNX='tas2es',xfuns='tas2es',plot=TRUE)
  dse.fw <- DSensemble(pca.fw,predictor=slp,pattern="psl_Amon_ens_",plot=TRUE)
  save(file='dse.mu.climatrans.rda',dse.mu,dse.fw)
} else load('dse.mu.climatrans.rda')
mu.stations<-as.station(dse.mu)
fw.stations<-as.station(dse.fw)

## Plot the downscaled results for the three megacities:
for (i in 1:3) {
  plot(mu.stations[[i]],new=FALSE)
  plot(fw.stations[[i]],new=FALSE)
}
```

@@

Check the long-term change: historic trend in the mean - include all seasons, but to get a clearer ipression, plot the anomalies (i.e. exclude the mean seasonal cycle).

```{r}
  load('climatrans.tx.rda')
  plot(as.4seasons(anomaly(climatrans.tx)))
```

Downscale the seasonal seasonal mean maximum temperature and then use the mean temperature to estimate the number of hot days assuming that the daily distribution is approximately Gaussian for a given season.

```{r}
  predictor <- retrieve('air.mon.mean.nc',param='air',lon=c(60,90),lat=c(10,35))

  trh <- c(40,30,32)
  for (i in 1:3) {
    y <- subset(climatrans.tx,is=i)
    if (!file.exists(paste('dse.tx.climatrans.',loc(y),'.rda',sep=''))) {
      dse.tx <- DSensemble.t2m(y,biascorrect=TRUE,type='ncdf4',
                           predictor=predictor,nmin=60,verbose=FALSE)
      save(file=paste('dse.tx.climatrans.',loc(y),'.rda',sep=''),dse.tx)
    } else load(paste('dse.tx.climatrans.',loc(y),'.rda',sep=''))

    ## The function hotsummerdays estimates the number of hot days based on
    ## the assumption that the temperature is close to being normally distributed
    hw <- hotsummerdays(x=y,dse=dse.tx,threshold=trh[i],plot=FALSE)
    plot(hw,new=FALSE)
  }
```

@@

Check the long-term change: historic trend in the seasonal spread (standard deviation) - the above estimates assumed that the spread is constant. The number of hot days will be even higher if the spread increases over time.

```{r}
  plot(as.4seasons(anomaly(climatrans.tx),FUN='sd'),new=FALSE)
```

### 2016-04-07 KSS/dr-pr4KSS.Rmd
Results for Climatrans: precipitation and hot summer days

```{r,eval=FALSE}
library(rmarkdown); render('climatrans.Rmd',pdf_document())
```

@@

@@

```{r}
## Downscaling analysis
  ds.mu.c <- DS(pca.mu.c,eof.es.c)
  plot(ds.mu.c)
```

We have now identified one connection between the large-scale conitions - vapour saturation pressure over the North Atlantic (a main moisture source) and the wet-day mean. There may be other fators affecting thestatistics too, and we can search for remaining structure in the data in terms of spatio-temporal variability. If the reminder looks onisy (e.g. 'flat' eigenvalues), then the remainder may be assumed to be from local effects and can be represented by a noise model.

```{r}
## Examine the PCA of residuals to see if there are any structures left
  mu.c.2 <- as.residual(as.station(ds.mu.c))
  mu.c.2 <- attrcp(mu.c,mu.c.2); class(mu.c.2) <- class(mu.c)
  pca.mu.c.2 <- PCA(mu.c.2)
  plot(pca.mu.c.2)
```

```{r}
## Downscaling analysis
  dse.mu.c <- DSensemble(pca.mu.c,predictor=es.c,
                         it=month.abb[c(1:3,10:12)],
                         FUNX='C.C.eq',plot=TRUE,verbose=FALSE)

>>>>

```{r,eval=FALSE}
library(rmarkdown); render('dr-pr4KSS.Rmd',pdf_document())
```

@@

```{r}
## Show maps of trend
map(mu.c,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-1,1,by=0.1),rev=TRUE))
title(expression(paste('Trend in ',mu,' (mm/day per decade) for cold months')))
map(mu.w,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-0.5,0.5,by=0.1),rev=TRUE))
title(expression(paste('Trend in ',mu,' (mm/day per decade) for warm months')))
map(100*fw.c,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-2,2,by=0.1),rev=TRUE))
title(expression(paste('Trend in ',f[w],' (% per decade) for cold months')))
map(100*fw.w,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-1,1,by=0.1),rev=TRUE))
title(expression(paste('Trend in ',f[w],' (mm/day per decade) for cold months')))
```

Variance of an exponentially distributed variable is $\mu^2$:

```{r}
## Show maps of trend
mu.c2 <- mu.c
coredata(mu.c2) <- coredata(mu.c)^2
map(mu.c2,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-25,25,by=1),rev=TRUE))
title(expression(paste('Trend in ',mu^2,' (variance per decade) for cold months')))
mu.w2 <- mu.w
coredata(mu.w2) <- coredata(mu.w)^2
map(mu.w2,FUN='trend',new=FALSE,colbar=list(pal='t2m',breaks=seq(-10,10,by=1),rev=TRUE))
title(expression(paste('Trend in ',mu^2,' (variance per decade) for warm months')))
```

@@

```{r}
## Downscaling analysis for the cold months
  ds.mu.c <- DS(pca.mu.c,eof.es.c)
  plot(ds.mu.c)
```

```{r}
## Downscaling analysis for the warm months
  ds.mu.w <- DS(pca.mu.w,eof.es.w)
  plot(ds.mu.w)
```

```{r}
## Downscaling analysis for the cold months
  ds.mu.C <- DS(pca.mu.C,eof.es.c)
  plot(ds.mu.C)
```

```{r}
## Downscaling analysis for the warm months
  ds.mu.W <- DS(pca.mu.W,eof.es.w)
  plot(ds.mu.W)
```

```{r}
## Downscaling analysis for the cold months
  ds.fw.c <- DS(pca.fw.c,eof.slp.c)
  plot(ds.mu.c)
```

```{r}
## Downscaling analysis for the warm months
  ds.fw.w <- DS(pca.fw.w,eof.slp.w)
  plot(ds.mu.w)
```

##**The residuals of the multiple regression analysis**

We have now identified one connection between the large-scale conitions - vapour saturation pressure over the North Atlantic (a main moisture source) and the wet-day mean. There may be other fators affecting thestatistics too, and we can search for remaining structure in the data in terms of spatio-temporal variability. If the reminder looks onisy (e.g. 'flat' eigenvalues), then the remainder may be assumed to be from local effects and can be represented by a noise model.dr

```{r}
## Examine the PCA of residuals for the cold monthsto see if there are any structures left
  mu.c.2 <- as.residual(as.station(ds.mu.c))
  mu.c.2 <- attrcp(mu.c,mu.c.2); class(mu.c.2) <- class(mu.c)
  pca.mu.c.2 <- PCA(mu.c.2)
  plot(pca.mu.c.2)
```

```{r}
## Examine the PCA of residuals for the warm months to see if there are any structures left
  mu.w.2 <- as.residual(as.station(ds.mu.w))
  mu.w.2 <- attrcp(mu.w,mu.w.2); class(mu.w.2) <- class(mu.w)
  pca.mu.w.2 <- PCA(mu.w.2)
  plot(pca.mu.w.2)
```

```{r}
## Examine the PCA of residuals for the cold monthsto see if there are any structures left
  fw.c.2 <- as.residual(as.station(ds.fw.c))
  fw.c.2 <- attrcp(fw.c,fw.c.2); class(fw.c.2) <- class(fw.c)
  pca.fw.c.2 <- PCA(fw.c.2)
  plot(pca.fw.c.2)
```

```{r}
## Examine the PCA of residuals for the warm months to see if there are any structures left
  fw.w.2 <- as.residual(as.station(ds.fw.w))
  fw.w.2 <- attrcp(fw.w,fw.w.2); class(fw.w.2) <- class(fw.w)
  pca.fw.w.2 <- PCA(fw.w.2)
  plot(pca.fw.w.2)
```

```{r}
## Examine the time series of the residuals
  plot.zoo(pca.mu.w.2,type='single')
```

```{r}
## Examine the ACF for the leading PC of the residuals
  plot(acf(pca.mu.w.2[,1]))
```

```{r}
## Examine the power spectrum for the leading PC of the residuals
  plot(spectrum(pca.mu.w.2[,1]))
```

```{r}
## Examine the distribution of the leading PC of the residuals
  plot(qqnorm(coredata(pca.mu.w.2[,1])))
  qqline(coredata(pca.mu.w.2[,1])))
```

```{r}
fndseres <- 'dse.nordic.precip.rda'
if (!file.exists(fndseres)) {
## Downscaling analysis
  dse.mu.c <- DSensemble(pca.mu.c,predictor=es.c,
                         it=month.abb[c(1:3,10:12)],
                         FUNX='C.C.eq',plot=TRUE,verbose=FALSE)
  dse.fw.c <- DSensemble(pca.fw.c,predictor=slp.c,
                         it=month.abb[c(1:3,10:12)],
                         FUNX='C.C.eq',plot=TRUE,verbose=FALSE)
  dse.mu.w <- DSensemble(pca.mu.w,predictor=es.w,
                         it=month.abb[4:9],
                         FUNX='C.C.eq',plot=TRUE,verbose=FALSE)
  dse.fw.w <- DSensemble(pca.fw.w,predictor=slp.w,
                         it=month.abb[4:9],
                         FUNX='C.C.eq',plot=TRUE,verbose=FALSE)
  save(file=fndseres,dse.mu.c,dse.fw.c,dse.mu.w,dse.fw.w)
} else load(fndseres)
